---
title: "Milestone 1: NY State Report Card"
format:
  html:
    embed-resources: true
    df-print: paged
editor:
  markdown:
    wrap: 80
---

For this milestone we're exploring the New York State School Report Card 2024-25
release. 

We want to use this dataset to look at three questions:

1.  How does money affect pupil performance at these schools?
2.  How have test scores trended over the last few years, and what does the
    increase/dropoff look like for affluent schools versus poorer schools? And
    how does it compare in math versus reading?
3.  In what ways are disadvantaged schools disadvantaged versus their peers?

With more data, we'd also like to look at how AI and social media lead to performance changes of students over time.


```{r setup}
#| message: false
#| warning: false
library(tidyverse)
library(ggrepel)
library(ggtext)
library(scales)
```

## Data loading

The database has around 30 tables. The ones that look useful for the 
questions are the math and ELA exam tables, the expenditures table, the
chronic absenteeism table, and the two teacher tables (inexperienced teachers
and out-of-certification teachers).

```{r}
#| message: false
math_exams = read_csv("data/Annual_EM_MATH.csv", show_col_types = FALSE)
ela_exams  = read_csv("data/Annual_EM_ELA.csv",  show_col_types = FALSE)
expenditures = read_csv("data/Expenditures_per_Pupil.csv", show_col_types = FALSE)
absenteeism = read_csv("data/Chronic_Absenteeism.csv", show_col_types = FALSE)
inexp_teachers = read_csv("data/Inexperienced_Teachers.csv", show_col_types = FALSE)
out_of_cert = read_csv("data/Out_of_Cert_Teachers.csv", show_col_types = FALSE)
```

The exam tables are pretty big (around 650k rows each) because every row is a
school × assessment × subgroup combination. The expenditure and teacher tables
are at the district level, which is one row per district per year, so they're
much smaller.

```{r}
math_exams
```

```{r}
expenditures
```

```{r}
glimpse(math_exams)
```

```{r}
glimpse(expenditures)
```

```{r}
glimpse(inexp_teachers)
```

```{r}
math_exams |> count(YEAR)
```

The dataset only has 2024 and 2025. So when we say trends, we really only have two data points to work with for
this milestone. For the final, we want to grab earlier SRC releases.

```{r}
math_exams |> count(SUBGROUP_NAME)
```

The exam tables already break each result down into a long list of subgroups
(Economically Disadvantaged, English Language Learner, race/ethnicity, gender,
homeless, foster care, etc.). That's helpful, because it already provides a definition
of disadvantaged.

```{r}
math_exams |> count(ASSESSMENT_NAME)
```

MATH3_8 is a roll-up summary, but the MEAN_SCORE column is empty
for those rows. The grade-level rows (MATH3 through MATH8) are the ones
that actually have scale scores. So we'll filter to those when we want a number
to compare.

ENTITY_CD column is a 12-digit code that identifies both
districts and schools. District codes end in 0000, while school codes don't. The
expenditures table mixes both kinds of rows. we'll filter to districts only
when joining because we only care about the aggregate

```{r}
expenditures |>
  filter(YEAR == 2024) |>
  count(district = grepl("0000$", ENTITY_CD))
```

Around 700 districts and 4500 schools.

## Goal 1: Does money buy performance?

Our first instinct was a simple scatter — do districts that spend more per pupil
score higher?

```{r goal1-data}
#| message: false

# District-level average math score (grades 3-8, all students, 2024)
district_math = math_exams |>
  filter(
    ASSESSMENT_NAME %in% c("MATH3","MATH4","MATH5","MATH6","MATH7","MATH8"),
    SUBGROUP_NAME == "All Students",
    YEAR == 2024,
    grepl("0000$", ENTITY_CD)
  ) |>
  group_by(ENTITY_CD, ENTITY_NAME) |>
  summarize(avg_math_score = mean(as.numeric(MEAN_SCORE), na.rm = TRUE),
            .groups = "drop")

# % economically disadvantaged per district
district_pct_ed = math_exams |>
  filter(
    ASSESSMENT_NAME %in% c("MATH3","MATH4","MATH5","MATH6","MATH7","MATH8"),
    SUBGROUP_NAME %in% c("All Students", "Economically Disadvantaged"),
    YEAR == 2024,
    grepl("0000$", ENTITY_CD)
  ) |>
  group_by(ENTITY_CD, SUBGROUP_NAME) |>
  summarize(n = sum(as.numeric(TOTAL_COUNT), na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = SUBGROUP_NAME, values_from = n) |>
  mutate(pct_ed = 100 * `Economically Disadvantaged` / `All Students`) |>
  select(ENTITY_CD, pct_ed)

# Per-pupil spending split by source (2024, districts >= 100 pupils)
district_money = expenditures |>
  filter(YEAR == 2024, grepl("0000$", ENTITY_CD), PUPIL_COUNT_TOT >= 100) |>
  select(ENTITY_CD,
         per_pupil_spending = PER_FED_STATE_LOCAL_EXP,
         federal_pp         = PER_FEDERAL_EXP,
         statelocal_pp      = PER_STATE_LOCAL_EXP)

# Full joined dataset
districts = district_math |>
  inner_join(district_money, by = "ENTITY_CD") |>
  inner_join(district_pct_ed, by = "ENTITY_CD") |>
  filter(!is.na(avg_math_score), !is.na(pct_ed), !is.na(per_pupil_spending),
         !grepl("^0{6}", ENTITY_CD))
```

### The puzzle: spending and scores move in the wrong direction

```{r goal1-plot1}
#| fig-width: 8
#| fig-height: 5.5


ggplot(districts, aes(x = per_pupil_spending, y = avg_math_score)) +
  geom_point(alpha = 0.45, size = 2.2, color = "steelblue") +
  scale_x_continuous(
    labels = label_dollar(scale = 1e-3, suffix = "k"),
    limits = c(NA, 90000)
  ) +
  labs(
    title    = "Spending doesn't predict scores — if anything, the pattern runs backwards",
    subtitle = "Each point is a NY school district, 2024–25.",
    x        = "Per-pupil spending (federal + state + local)",
    y        = "Average grade 3–8 math scale score",
    caption  = "Source: NY State School Report Card 2024–25 release. Districts with ≥ 100 pupils."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40", size = 10.5),
    plot.caption  = element_text(color = "gray55", size = 8)
  )
```

There's no upward pattern here — if anything, higher spending is weakly
associated with *lower* scores, the opposite of what we expected. So what's
going on? The next chart shows why.

### Why high-need districts spend more: equity funding at work

```{r goal1-plot2}
#| fig-width: 8
#| fig-height: 4.8

districts_q = districts |>
  mutate(need_group = cut(
    pct_ed,
    breaks = quantile(pct_ed, probs = 0:5/5, na.rm = TRUE),
    labels = c("Lowest need\n(0–20th %ile)", "Low need\n(20–40th %ile)",
               "Middle\n(40–60th %ile)",     "High need\n(60–80th %ile)",
               "Highest need\n(80–100th %ile)"),
    include.lowest = TRUE
  ))

funding_long = districts_q |>
  group_by(need_group) |>
  summarize(Federal         = mean(federal_pp,    na.rm = TRUE),
            `State & Local` = mean(statelocal_pp, na.rm = TRUE),
            .groups = "drop") |>
  pivot_longer(cols = c("Federal", "State & Local"),
               names_to = "source", values_to = "spending") |>
  mutate(source = factor(source, levels = c("State & Local", "Federal")))

totals = funding_long |>
  group_by(need_group) |>
  summarize(total = sum(spending), .groups = "drop")

ggplot(funding_long, aes(x = need_group, y = spending, fill = source)) +
  geom_col(width = 0.65) +
  geom_text(data = totals,
            aes(x = need_group, y = total,
                label = dollar(total, scale = 1e-3, suffix = "k")),
            inherit.aes = FALSE,
            vjust = -0.4, size = 3.4, fontface = "bold", color = "gray20") +
  scale_y_continuous(labels = label_dollar(scale = 1e-3, suffix = "k"),
                     expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("Federal" = "#e08214", "State & Local" = "#4d9de0"),
                    name = "Funding source") +
  labs(
    title    = "Equity aid flows to high-need districts — but state & local funding barely changes",
    subtitle = "Average per-pupil spending by district needs quintile, 2024–25.",
    x        = NULL, y = "Average per-pupil spending",
    caption  = "Source: NY State School Report Card 2024–25 release."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 13.5),
    plot.subtitle = element_text(color = "gray40", size = 10.5),
    legend.position = "top",
    plot.caption  = element_text(color = "gray55", size = 8),
    panel.grid.major.x = element_blank()
  )
```

High-need districts do receive more money overall, mainly through federal aid.
But that extra funding hasn't translated into higher scores — which suggests
money alone isn't enough to close the gap.

### The real driver: poverty, not spending

```{r goal1-plot3}
#| fig-width: 8
#| fig-height: 5.5

ggplot(districts, aes(x = pct_ed, y = avg_math_score, color = per_pupil_spending)) +
  geom_point(alpha = 0.55, size = 2.2) +
  geom_smooth(method = "lm", color = "gray20", linewidth = 1,
              se = FALSE) +
  scale_x_continuous(labels = function(x) paste0(x, "%")) +
  scale_color_viridis_c(
    option = "plasma", direction = -1,
    name   = "Per-pupil\nspending",
    labels = label_dollar(scale = 1e-3, suffix = "k"),
    breaks = c(20000, 35000, 50000, 75000)
  ) +
  labs(
    title    = "Poverty rate predicts scores far more cleanly than spending does",
    subtitle = "Higher-spending districts (purple) cluster at the high-poverty end.",
    x        = "% of students economically disadvantaged",
    y        = "Average grade 3–8 math scale score",
    caption  = "Source: NY State School Report Card 2024–25 release."
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 13.5),
    plot.subtitle = element_text(color = "gray40", size = 10.5),
    legend.position = "right",
    plot.caption  = element_text(color = "gray55", size = 8)
  )
```

```{r goal1-cors}
#| echo: false
r_spend = cor(districts$per_pupil_spending, districts$avg_math_score,
              use = "complete.obs") |> round(2)
r_pov   = cor(districts$pct_ed, districts$avg_math_score,
              use = "complete.obs") |> round(2)
```

Poverty rate is a strong predictor of scores (r = `r r_pov`), far more so than
spending (r = `r r_spend`). That raises the next question: how wide is this
gap, and is it closing?

## Goal 2: Affluent vs poorer schools in math and reading

Since the dataset already labels each result as Economically Disadvantaged or
Not Economically Disadvantaged, we can compare those two groups directly.

```{r}
math_by_subgroup = math_exams |>
  filter(ASSESSMENT_NAME %in% c("MATH3","MATH4","MATH5","MATH6","MATH7","MATH8"),
         SUBGROUP_NAME %in% c("Economically Disadvantaged",
                              "Not Economically Disadvantaged")) |>
  group_by(YEAR, SUBGROUP_NAME) |>
  summarize(avg_score = mean(as.numeric(MEAN_SCORE), na.rm = TRUE),
            .groups = "drop") |>
  mutate(subject = "Math")

ela_by_subgroup = ela_exams |>
  filter(ASSESSMENT_NAME %in% c("ELA3","ELA4","ELA5","ELA6","ELA7","ELA8"),
         SUBGROUP_NAME %in% c("Economically Disadvantaged",
                              "Not Economically Disadvantaged")) |>
  group_by(YEAR, SUBGROUP_NAME) |>
  summarize(avg_score = mean(as.numeric(MEAN_SCORE), na.rm = TRUE),
            .groups = "drop") |>
  mutate(subject = "ELA (Reading)")

scores_by_subgroup = rbind(math_by_subgroup, ela_by_subgroup)

scores_by_subgroup |>
  ggplot(aes(x = factor(YEAR), y = avg_score,
             color = SUBGROUP_NAME, group = SUBGROUP_NAME)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~ subject) +
  labs(
    x = "Year",
    y = "Average scale score",
    color = NULL,
    title = "The gap between disadvantaged and not is wide and not closing",
    subtitle = "Average grade 3-8 scale score, NY State"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
scores_by_subgroup
```

The gap is real big about 13-14 scale score points in both subjects and it
basically doesn't move from 2024 to 2025. The math gap looks slightly bigger
than the ELA gap, which is think is interesting but we'd want to double-check
that with a more careful comparison since math and ELA scales aren't
necessarily on the same units.

## Goal 3: How are disadvantaged schools disadvantaged?

The score gap shows up in Goal 2, but we wanted to know whether the gap has
structural causes too, things like teacher quality and absenteeism. The
chronic absenteeism table also has the same subgroup breakdown, so we can do a
direct comparison there.

```{r}
absent_2024 = absenteeism |>
  filter(YEAR == 2024) |>
  mutate(ENROLLMENT   = suppressWarnings(as.numeric(ENROLLMENT)),
         ABSENT_COUNT = suppressWarnings(as.numeric(ABSENT_COUNT)),
         ABSENT_RATE  = suppressWarnings(as.numeric(ABSENT_RATE)))

# Chronic_Absenteeism doesn't report "Not Economically Disadvantaged" directly,
# so back-calculate it at the school level: All Students minus Econ. Disadvantaged.
not_ed = absent_2024 |>
  filter(SUBGROUP_NAME %in% c("All Students", "Economically Disadvantaged")) |>
  select(ENTITY_CD, SUBGROUP_NAME, ENROLLMENT, ABSENT_COUNT) |>
  pivot_wider(names_from = SUBGROUP_NAME,
              values_from = c(ENROLLMENT, ABSENT_COUNT)) |>
  mutate(enroll = `ENROLLMENT_All Students` - `ENROLLMENT_Economically Disadvantaged`,
         absent = `ABSENT_COUNT_All Students` - `ABSENT_COUNT_Economically Disadvantaged`,
         ABSENT_RATE = ifelse(enroll > 0, 100 * absent / enroll, NA)) |>
  filter(!is.na(ABSENT_RATE)) |>
  summarize(SUBGROUP_NAME = "Not Economically Disadvantaged",
            avg_absent_rate = mean(ABSENT_RATE, na.rm = TRUE))

absent_by_group = absent_2024 |>
  filter(SUBGROUP_NAME %in% c("Economically Disadvantaged",
                              "English Language Learner",
                              "Students with Disabilities",
                              "All Students")) |>
  group_by(SUBGROUP_NAME) |>
  summarize(avg_absent_rate = mean(ABSENT_RATE, na.rm = TRUE),
            .groups = "drop") |>
  bind_rows(not_ed)

absent_by_group |>
  ggplot(aes(x = reorder(SUBGROUP_NAME, avg_absent_rate),
             y = avg_absent_rate,
             fill = SUBGROUP_NAME)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", avg_absent_rate)), hjust = -0.2) +
  coord_flip() +
  ylim(0, 35) +
  labs(
    x = NULL,
    y = "Average chronic absenteeism rate (%)",
    title = "Disadvantaged students miss a lot more school",
    subtitle = "Chronic absenteeism = missing 10% or more of school days, 2024"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The disadvantaged group's chronic absenteeism rate is roughly double the
not-disadvantaged group. That feels like a big deal because if you're not in
the building you can't catch up academically, so this might be one
mechanism behind the score gap.

The teacher tables do not include subgroup breakdowns, so we compare teacher staffing at the district level instead. We group districts by their share of economically disadvantaged students, then look at whether high-need districts have higher percentages of inexperienced teachers and teachers working out of certification.

```{r}
district_ed = math_exams |>
  mutate(
    ENTITY_CD = as.character(ENTITY_CD),
    TOTAL_COUNT = as.numeric(TOTAL_COUNT)
  ) |>
  filter(
    YEAR == 2024,
    ASSESSMENT_NAME %in% c("MATH3", "MATH4", "MATH5", "MATH6", "MATH7", "MATH8"),
    SUBGROUP_NAME %in% c("All Students", "Economically Disadvantaged"),
    grepl("0000$", ENTITY_CD)
  ) |>
  group_by(ENTITY_CD, ENTITY_NAME, SUBGROUP_NAME) |>
  summarize(total_students = sum(TOTAL_COUNT, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(
    names_from = SUBGROUP_NAME,
    values_from = total_students
  ) |>
  mutate(
    pct_econ_disadvantaged =
      100 * `Economically Disadvantaged` / `All Students`
  ) |>
  filter(!is.na(pct_econ_disadvantaged)) |>
  mutate(
    need_group_num = ntile(pct_econ_disadvantaged, 4),
    need_group = case_when(
      need_group_num == 1 ~ "Low-need districts\n(bottom 25%)",
      need_group_num %in% c(2, 3) ~ "Middle districts",
      need_group_num == 4 ~ "High-need districts\n(top 25%)"
    ),
    need_group = factor(
      need_group,
      levels = c(
        "Low-need districts\n(bottom 25%)",
        "Middle districts",
        "High-need districts\n(top 25%)"
      )
    )
  )

teacher_dists = bind_rows(
  inexp_teachers |>
    mutate(ENTITY_CD = as.character(ENTITY_CD)) |>
    filter(YEAR == 2024, grepl("0000$", ENTITY_CD)) |>
    transmute(
      ENTITY_CD,
      ENTITY_NAME,
      pct = as.numeric(PER_TEACH_INEXP),
      metric = "Inexperienced teachers (<4 yrs)"
    ),

  out_of_cert |>
    mutate(ENTITY_CD = as.character(ENTITY_CD)) |>
    filter(YEAR == 2024, grepl("0000$", ENTITY_CD)) |>
    transmute(
      ENTITY_CD,
      ENTITY_NAME,
      pct = as.numeric(PER_OUT_CERT),
      metric = "Teaching out of certification"
    )
)

teacher_need = teacher_dists |>
  left_join(
    district_ed |> select(ENTITY_CD, pct_econ_disadvantaged, need_group),
    by = "ENTITY_CD"
  ) |>
  filter(!is.na(pct), !is.na(pct_econ_disadvantaged))

teacher_need |>
  ggplot(aes(x = need_group, y = pct, fill = need_group)) +
  geom_boxplot(width = 0.5, alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1.6) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    x = NULL,
    y = "Teachers in category (%)",
    title = "High-need districts tend to have more teacher staffing instability",
    subtitle = "Districts are grouped by their share of economically disadvantaged students, 2024",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The chart shows that high-need districts generally have higher percentages of inexperienced teachers and teachers working out of certification. Each jittered point represents one district, while the boxplots summarize the overall distribution within each need group. The pattern is not identical for every district, but the distribution shifts upward for high-need districts, especially in the right tail. This suggests that disadvantaged districts may face structural challenges beyond student test scores, including less stable teacher staffing.