---
title: "Milestone 1: NY State Report Card"
format:
  html:
    embed-resources: true
    df-print: paged
editor:
  markdown:
    wrap: 80
---

For this milestone we're exploring the New York State School Report Card 2024-25
release. 

We want to use this dataset to look at three questions:

1.  How does money affect pupil performance at these schools?
2.  How have test scores trended over the last few years, and what does the
    increase/dropoff look like for affluent schools versus poorer schools? And
    how does it compare in math versus reading?
3.  In what ways are disadvantaged schools disadvantaged versus their peers?

With more data, we'd also like to look at how AI and social media lead to performance changes of students over time.


```{r setup}
#| message: false
#| warning: false
library(tidyverse)
library(ggrepel)
```

## Data loading

The database has around 30 tables. The ones that look useful for the 
questions are the math and ELA exam tables, the expenditures table, the
chronic absenteeism table, and the two teacher tables (inexperienced teachers
and out-of-certification teachers).

```{r}
#| message: false
math_exams = read_csv("data/Annual_EM_MATH.csv", show_col_types = FALSE)
ela_exams  = read_csv("data/Annual_EM_ELA.csv",  show_col_types = FALSE)
expenditures = read_csv("data/Expenditures_per_Pupil.csv", show_col_types = FALSE)
absenteeism = read_csv("data/Chronic_Absenteeism.csv", show_col_types = FALSE)
inexp_teachers = read_csv("data/Inexperienced_Teachers.csv", show_col_types = FALSE)
out_of_cert = read_csv("data/Out_of_Cert_Teachers.csv", show_col_types = FALSE)
```

The exam tables are pretty big (around 650k rows each) because every row is a
school × assessment × subgroup combination. The expenditure and teacher tables
are at the district level, which is one row per district per year, so they're
much smaller.

```{r}
math_exams
```

```{r}
expenditures
```

```{r}
glimpse(math_exams)
```

```{r}
glimpse(expenditures)
```

```{r}
glimpse(inexp_teachers)
```

```{r}
math_exams |> count(YEAR)
```

The dataset only has 2024 and 2025. So when we say trends, we really only have two data points to work with for
this milestone. For the final, we want to grab earlier SRC releases.

```{r}
math_exams |> count(SUBGROUP_NAME)
```

The exam tables already break each result down into a long list of subgroups
(Economically Disadvantaged, English Language Learner, race/ethnicity, gender,
homeless, foster care, etc.). That's helpful, because it already provides a definition
of disadvantaged.

```{r}
math_exams |> count(ASSESSMENT_NAME)
```

MATH3_8 is a roll-up summary, but the MEAN_SCORE column is empty
for those rows. The grade-level rows (MATH3 through MATH8) are the ones
that actually have scale scores. So we'll filter to those when we want a number
to compare.

ENTITY_CD column is a 12-digit code that identifies both
districts and schools. District codes end in 0000, while school codes don't. The
expenditures table mixes both kinds of rows. we'll filter to districts only
when joining because we only care about the aggregate

```{r}
expenditures |>
  filter(YEAR == 2024) |>
  count(district = grepl("0000$", ENTITY_CD))
```

Around 700 districts and 4500 schools.

## Goal 1: Does money buy performance?

Take each district's per-pupil spending and its average math score, and plot one against the other.

```{r}
district_math = math_exams |>
  filter(ASSESSMENT_NAME %in% c("MATH3","MATH4","MATH5","MATH6","MATH7","MATH8"),
         SUBGROUP_NAME == "All Students",
         YEAR == 2024,
         grepl("0000$", ENTITY_CD)) |>
  group_by(ENTITY_CD, ENTITY_NAME) |>
  summarize(avg_math_score = mean(as.numeric(MEAN_SCORE), na.rm = TRUE),
            .groups = "drop")

district_money = expenditures |>
  filter(YEAR == 2024,
         grepl("0000$", ENTITY_CD),
         PUPIL_COUNT_TOT >= 100) |>
  select(ENTITY_CD, per_pupil_spending = PER_FED_STATE_LOCAL_EXP)

money_vs_score = merge(district_math, district_money, by = "ENTITY_CD")

money_vs_score |>
  ggplot(aes(x = per_pupil_spending, y = avg_math_score)) +
  geom_point(alpha = 0.35, color = "steelblue") +
  geom_smooth(method = "lm", color = "firebrick", se = FALSE) +
  labs(
    x = "Per-pupil spending ($, 2024)",
    y = "Average grade 3-8 math scale score",
    title = "Money doesn't seem to be doing the work",
    subtitle = "Each point is a NY district"
  ) +
  theme_minimal()
```

```{r}
cor(money_vs_score$per_pupil_spending, money_vs_score$avg_math_score)
```

We expected the line to slope up and it kind of does the opposite, slightly. Our
guess is that high-need districts already get more per-pupil funding through
state and federal equity aid, so the districts spending the most aren't
actually the affluent ones. That makes spending a weird axis to look at on its
own, and pushes us toward Goal 2 where we look at affluence directly.

## Goal 2: Affluent vs poorer schools in math and reading

Since the dataset already labels each result as Economically Disadvantaged or
Not Economically Disadvantaged, we can compare those two groups directly.

```{r}
math_by_subgroup = math_exams |>
  filter(ASSESSMENT_NAME %in% c("MATH3","MATH4","MATH5","MATH6","MATH7","MATH8"),
         SUBGROUP_NAME %in% c("Economically Disadvantaged",
                              "Not Economically Disadvantaged")) |>
  group_by(YEAR, SUBGROUP_NAME) |>
  summarize(avg_score = mean(as.numeric(MEAN_SCORE), na.rm = TRUE),
            .groups = "drop") |>
  mutate(subject = "Math")

ela_by_subgroup = ela_exams |>
  filter(ASSESSMENT_NAME %in% c("ELA3","ELA4","ELA5","ELA6","ELA7","ELA8"),
         SUBGROUP_NAME %in% c("Economically Disadvantaged",
                              "Not Economically Disadvantaged")) |>
  group_by(YEAR, SUBGROUP_NAME) |>
  summarize(avg_score = mean(as.numeric(MEAN_SCORE), na.rm = TRUE),
            .groups = "drop") |>
  mutate(subject = "ELA (Reading)")

scores_by_subgroup = rbind(math_by_subgroup, ela_by_subgroup)

scores_by_subgroup |>
  ggplot(aes(x = factor(YEAR), y = avg_score,
             color = SUBGROUP_NAME, group = SUBGROUP_NAME)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  facet_wrap(~ subject) +
  labs(
    x = "Year",
    y = "Average scale score",
    color = NULL,
    title = "The gap between disadvantaged and not is wide and not closing",
    subtitle = "Average grade 3-8 scale score, NY State"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
scores_by_subgroup
```

The gap is real big about 13-14 scale score points in both subjects and it
basically doesn't move from 2024 to 2025. The math gap looks slightly bigger
than the ELA gap, which is think is interesting but we'd want to double-check
that with a more careful comparison since math and ELA scales aren't
necessarily on the same units.

## Goal 3: How are disadvantaged schools disadvantaged?

The score gap shows up in Goal 2, but we wanted to know whether the gap has
structural causes too, things like teacher quality and absenteeism. The
chronic absenteeism table also has the same subgroup breakdown, so we can do a
direct comparison there.

```{r}
absent_2024 = absenteeism |>
  filter(YEAR == 2024) |>
  mutate(ENROLLMENT   = suppressWarnings(as.numeric(ENROLLMENT)),
         ABSENT_COUNT = suppressWarnings(as.numeric(ABSENT_COUNT)),
         ABSENT_RATE  = suppressWarnings(as.numeric(ABSENT_RATE)))

# Chronic_Absenteeism doesn't report "Not Economically Disadvantaged" directly,
# so back-calculate it at the school level: All Students minus Econ. Disadvantaged.
not_ed = absent_2024 |>
  filter(SUBGROUP_NAME %in% c("All Students", "Economically Disadvantaged")) |>
  select(ENTITY_CD, SUBGROUP_NAME, ENROLLMENT, ABSENT_COUNT) |>
  pivot_wider(names_from = SUBGROUP_NAME,
              values_from = c(ENROLLMENT, ABSENT_COUNT)) |>
  mutate(enroll = `ENROLLMENT_All Students` - `ENROLLMENT_Economically Disadvantaged`,
         absent = `ABSENT_COUNT_All Students` - `ABSENT_COUNT_Economically Disadvantaged`,
         ABSENT_RATE = ifelse(enroll > 0, 100 * absent / enroll, NA)) |>
  filter(!is.na(ABSENT_RATE)) |>
  summarize(SUBGROUP_NAME = "Not Economically Disadvantaged",
            avg_absent_rate = mean(ABSENT_RATE, na.rm = TRUE))

absent_by_group = absent_2024 |>
  filter(SUBGROUP_NAME %in% c("Economically Disadvantaged",
                              "English Language Learner",
                              "Students with Disabilities",
                              "All Students")) |>
  group_by(SUBGROUP_NAME) |>
  summarize(avg_absent_rate = mean(ABSENT_RATE, na.rm = TRUE),
            .groups = "drop") |>
  bind_rows(not_ed)

absent_by_group |>
  ggplot(aes(x = reorder(SUBGROUP_NAME, avg_absent_rate),
             y = avg_absent_rate,
             fill = SUBGROUP_NAME)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", avg_absent_rate)), hjust = -0.2) +
  coord_flip() +
  ylim(0, 35) +
  labs(
    x = NULL,
    y = "Average chronic absenteeism rate (%)",
    title = "Disadvantaged students miss a lot more school",
    subtitle = "Chronic absenteeism = missing 10% or more of school days, 2024"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

The disadvantaged group's chronic absenteeism rate is roughly double the
not-disadvantaged group. That feels like a big deal because if you're not in
the building you can't catch up academically, so this might be one
mechanism behind the score gap.

The teacher tables don't have subgroup breakdowns, so for now we're looking
at how the percent of inexperienced teachers and out-of-certification teachers
distribute across districts.

```{r}
teacher_dists = inexp_teachers |>
  filter(YEAR == 2024) |>
  select(ENTITY_CD, pct = PER_TEACH_INEXP) |>
  mutate(metric = "Inexperienced teachers (<4 yrs)") |>
  rbind(
    out_of_cert |>
      filter(YEAR == 2024) |>
      select(ENTITY_CD, pct = PER_OUT_CERT) |>
      mutate(metric = "Teaching out of certification")
  )

teacher_dists |>
  ggplot(aes(x = as.numeric(pct), fill = metric)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    x = "Percent of teachers in district",
    y = "Number of districts",
    title = "Distribution of inexperienced and out-of-certification teachers across NY districts",
    subtitle = "Each district contributes one observation per metric, 2024"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Most districts are clustered at the low end but there's a notable
right tail, with some districts where 30%+ of teachers are
inexperienced or 20%+ are teaching out of certification. We can merge this with each district's percent
economically disadvantaged, so we can answer are the districts in that right
tail also the high-need districts?